DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems

### 2012-11-23 Paper_May_2012/main.tex
Todos eingefuegt (aber noch nicht style file). Wollte einmal speichern bevor ich alles im neuen style file umkopiere

\newcommand{\todo}[1]{\textbf{\textsc{\textcolor{black}{TODO: #1}}}}
%\newcommand{\mymarginpar}[1]{\marginpar{\textcolor{red}{\scriptsize{#1}}}}
\newcommand{\mymarginpar}[1]{\marginpar{\textcolor{red}{\normalsize{#1}}}}
%
\begin{document}
\markboth{C. Goll, T. Wick, and W. Wollner}{DOpElib:  A Goal Oriented Software Library}

@@

\begin{abstract}
In this article, we describe the software library 
\textit{Differential Equations and Optimization Environment} (\dope{}).
The main feature of \dope{} is that it provides a unified interface to high level algorithms 
such as time-stepping methods, nonlinear solvers and optimization routines. This structure ensures 
that first of all the user is required to write only those sections of code that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with only a few lines of code to change instead of large reimplementations.
The article illustrates the design principles and various features
of the \dope{} and provides some 
numerical results as demonstration for the versatility of the software.
\end{abstract}

\begin{bottomstuff}
Author's adresses: C. Goll, Institut f\"ur Angewandte Mathematik,
Universit\"at Heidelberg;
T. Wick, The Institute for Computational Engineering and Sciences;
W. Wollner, Fachbereich Mathematik, Universit\"at Hamburg.
\end{bottomstuff}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exist a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \cite{dealnew}, dune \cite{dune}, 
FEniCS \cite{fenics}
or commercial solvers as for instance ANSYS (with its Fluent package),
COMSOL, or simufact forming.
Finally, often, a lot of work is (still) done in Matlab.
In fact, these libraries provide very useful tools 
for a huge variety of applications.

However, often, the user is confronted with 
certain bottlenecks. On the one hand, 
in open-source projects for academic research, 
standard high-level routines 
which are necessary to solve even simplest PDEs,
such as linear and nonlinear solvers or time-stepping schemes must be implemented 
by the users themselves. However, such routines are, at large,
problem-independent and could be provided as standard tools. 
As our guiding design principle, we note that in many 
numerical studies, the main task is the computation of some  
target functional, e.g., error norms,
point values, or certain mean-values as drag or lift coefficients.
As a second motivating principle, 
there is nowadays an increasing interest in computing 
optimization problems in which the state equation is governed 
by a PDE. Such optimization problems comprise optimal control,
parameter estimation, inverse problems or optimal experimental design. 
To do this in complex applications usually requires tremendous work already
on the PDE itself before considering optimization of the processes modeled 
by the PDE. Thus it is advantageous to have an environment that allows to 
focus on the development of an appropriate code for the PDE at first but 
asserts due to the required interface, that the code written for the PDE
is then reusable for the application of state-of-the-art optimization 
routines.
So far, there exist a very limited number of open source packages
which fulfill the requirements listed above; 
sucessful in-house code examples are Gascoigne \cite{gascoigne}
and RoDoBo \cite{rodobo}. 
Both have been initiated as well in 
the Numerical Analysis Group in Heidelberg.

@@

The presented toolkit \dope{}: 
\textit{Differential Equations and Optimization Environment} \cite{dope}
fills this gap. We present a software
to solve both PDE problems as well as optimization problems 
constrained by a PDE. 
Its main feature is to give a unified interface to high level algorithms such as 
time-stepping methods, nonlinear solvers and optimization routines. 
\dope{} is designed such that the user only needs to write those parts
of the code that are problem dependent while all invariant 
parts of the algorithms
are reusable without any need for further coding.
In particular, the user is enabled to switch between various different 
algorithms without the need to rewrite the problem dependent code, 
though obviously he or she will
have to replace the algorithm object with an other one. 
This replacement can be done by replacing the appropriate object at only
one point in the code.

@@

The project was initially initiated at the University of Heidelberg in 2010 and  
at the present stage the following key features are supported by the library
\begin{itemize}
\item Solution of nonlinear stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various, easily exchangeable, 
  time-stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item Use of all finite elements of from \deal{} including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs by elimination of 
   the state variable.
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other control and state constraints.
\item Several examples showing how to solve various kinds of optimization 
  problems involving PDE constraints.
\item Goal-oriented mesh adaptation with the Dual Weighted Residual method and
  standard residual based error estimators.
\item Different spatial triangulations for control and state variables without
  the need to reimplement the user dependent code when switching from one common 
  triangulation to different ones.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
This library is designed to allow easy implementation and numerical solutions 
of problems involving partial differential equations (PDEs). 
In particular, we remove all tedious routine implementation needed in 
\deal{} by introducing appropriate classes that take care of all the 
problem independent implementation while leaving only the problem specific 
part to the user. To show how this is designed we will, in the sequel, 
describe the reasoning first for a simple stationary PDE in 
Section~\ref{subsubsec:stationary problems} and then show
how this idea extends to non stationary problems in 
Section~\ref{sec:timedep} as well as to optimization
problems with PDE constraints in Section~\ref{sec:opt}.

\subsection{Solving of Stationary PDEs}
\label{subsubsec:stationary problems}
For the start we consider a prototypical problem given by the solution of 
a stationary PDE, which we assume to be given in weak form.
In abstract notation it reads; find some $u$ in an appropriate set
such that
\begin{align}\label{eq:prototype_weak}
a(u)(\phi) = (f,\phi) \quad \forall \phi \in V,
\end{align}
with some suitable space $V$ for the test functions and a 
right hand side function $f$.
To illustrate this further, we take the stationary 
incompressible Navier-Stokes equations in a domain $\Omega \subset \R^d$, 
i.e., $u = (v,p)$ and $\phi = (\phi_u,\phi_p)$ then
\begin{align}\label{eq:ns}
a(u)(\phi) = \nu(\nabla v, \nabla\phi_u) + (v \nabla v,\phi_u) + (p, \nabla \cdot \phi_u) + (\nabla \cdot v ,\phi_p)
\end{align}
for all $\phi \in V = H^1_0(\Omega;\R^d) \times L^2_0(\Omega)$.
Here $(\cdot,\cdot)$ denotes the $L^2$-inner-product.
The solution will be searched for in $V + (u_d,0)$ where $u_d$ is some 
continuation of the given Dirichlet boundary data.
Then typically we do not only want to calculate $u$ but more likely we are 
interested in one (or several) functionals depending on the solution, i.e., some numbers $J(u)$.

In order to solve this problem, i.e., calculate $J(u)$, we need to discretize 
the equation. In \dope, we are using finite elements to do this. This means we need to pick a 
subdivision $\mathcal T_h$, for instance into quadrilaterals,
of the domain $\Omega$ and a finite element. In the concrete case, we could, 
for instance, take the $\mathcal Q_2/\mathcal Q_1$ Taylor-Hood element.
As it is clear, that the initialization of the unknowns, the local shape 
functions, removal of hanging node constraints, etc. always follows the 
same procedure, we define an object which takes the finite element (FE)
and the triangulation and takes care of all other things related to
handling the finite element space $V_h$. We will refer, to 
this as the \texttt{SpaceTimeHandler}. Then, it is clear that the simplest 
form of the constructor should have the following interface:
\begin{lstlisting}
  template<typename Triangulation, typename FE>
      void SpaceTimeHandler(Triangulation &, FE);
\end{lstlisting}
Next, one needs to solve the discrete equation
\begin{align}\label{eq:discrete_equation}
a(u_h)(\phi_h) = (f,\phi_h) \quad \forall \phi \in V_h.
\end{align}
To this end, one can employ a newton type method with some globalization 
like line-search. To do so, one needs to calculate the residual of the 
equation, as well as to solve linearized equations for the calculation of 
the newton method and then employing some kind of globalization strategy.
This again will usually require the evaluation of the residual.

@@

\begin{remark}
In fact, the above constructors are not exactly what is implemented in \dope{}.
This is due to the fact, that for instance 
there are several different sparsity patterns available in 
\deal{} which the user should be able chose. Thus it needs to be 
given to the \texttt{SpaceTimeHandler}. However, to keep the 
presentation clear we refrained from including these details above and in
what follows.
\end{remark}

In the next section we will now sketch the way a user needs to implement the
user-dependent data as the local cell integrals and the initialization of the
objects described above.
\subsubsection{Problem dependent implementation of stationary PDE problems}
To solve a stationary PDE problem, the user has to provide some code which can roughly be divided into the following main sections:
\begin{description}
\item[Discretization] Specify what kind of triangulation,
finite element, quadrature rules, etc. we want to use.
\item[Problem-data] Define the weak form, the corresponding matrices, the functionals as well as the boundary conditions. With exception of the boundary conditions, we have to specify all this only cell-wise, \dope{} handles the rest for us.
\item[Solvers] Which (non-)linear solver should we use? 
\item[Additional features] ParameterReader to read in the run-time parameters
  such as material coefficients or adjustments for the solvers like tolerance
  for Newton's method. Second, definition of Outputhandler etc.
\end{description}

We describe now a program designed to compute a value $J(u)$, where $u=(v,p)$ is the solution of the Navier-Stokes equations \eqref{eq:ns} on the domain $\Omega$ as seen in figure~\ref{fig:example_ns} with  $f=0$, and $J$ is the drag-force of the enclosed circle. We assume given Dirichlet data for the velocity $v$ on the boundary $\Gamma_{D}=\Gamma_{in}\cup\Gamma_{wall}\cup\Gamma_{circ}$. This is the Benchmark-Case 2D-1 from \cite{TuSchae96}, see there for further details.
\begin{figure}[h]
\centering
\resizebox{0.5\textwidth}{!}{\input{Pictures/example_ns}}
\caption{Flow around a cylinder with 
circle-center $C=(0.2,0.2)$ and radius $r=0.05$.}
\label{fig:example_ns}
\end{figure}

We split the program into the parts \texttt{main.cc}, \texttt{functionals.h} and \texttt{localpde.h}. The \texttt{main.cc}-file is responsible for the declaration of various objects which have mostly to do with the \textit{discretization}, the \textit{solvers} and the general \textit{program sequence}, whereas the other two contain most of the problem-data -- especially the cell-wise contributions of the weak form, the right hand side and the functional. We give a brief explanation on every file in the following.

@@

\end{lstlisting}
Next, we incorporate the Dirichlet data, which is given by the class \texttt{DirichleData}. Then we have to tell the \texttt{prob\_container} for which components and boundaries we prescribe the Dirichlet conditions. This is done via a boundary-color (0 stands for $\Gamma_D$) and a \texttt{std::vector<bool> comp\_mask = (true, true, false)}.
\begin{lstlisting}

@@

  typedef StatPDEProblem<NEWTONSOLVER> SSOLVER;

\end{lstlisting}
Here, \texttt{NEWTONSOLVER} itself is an abbreviation of
\begin{lstlisting}

  typedef NewtonSolver<LINEARSOLVER> NEWTONSOLVER;

\end{lstlisting}
The  template parameter \texttt{LINEARSOLVER} describes which linear solver we use in this example. We opted for a direct solver, so: 
\begin{lstlisting}

  typedef DirectLinearSolver<MATRIX, VECTOR> LINEARSOLVER;

@@

  SSOLVER solver(prob_container, pr,
                    quad_formula, face_quad_formula);

\end{lstlisting}
Finally, the program is started by
\begin{lstlisting}

  solver.ReInit();
  solver.ComputeReducedFunctionals();

@@

Let $N$ be the number of degrees of freedom of our discretization and $\Set{\phi_{h}^{i}|1\leq i \leq N}$ be a given FE-Basis. As stated in subsection~\ref{subsubsec:stationary problems}, to solve the discrete equation \eqref{eq:discrete_equation} we need the residual, i.e. the vector
\begin{align}\label{eq:discrete_residual}
\left(a(u_h)(\phi_{h}^{i})-(f,\phi_{h}^{i})\right)_{i=1}^N,
\end{align}
and the Jacobi-matrix of the weak form, i.e.
\begin{align}\label{eq:discrete_matrix}
\left(\nabla a(u_h)(\phi_{h,j},\phi_{h}^{i})\right)_{i,j=1}^N.
\end{align}
Now all we have to supply here are the cell-wise contributions to these terms, the \texttt{Integrator} takes care of assembling \eqref{eq:discrete_residual} and \eqref{eq:discrete_matrix}.

To this end, we define a class \texttt{LocalPDE} which provides the three methods
\texttt{CellEquation}, \texttt{CellRightHandSide} and \texttt{CellMatrix}. The implementation of the terms works by and large as in \deal{}. We show the basic structure of these methods using the example of \texttt{CellEquation}.

\begin{lstlisting}
  void CellEquation(const CellDataContainer<2>& cdc,
                    VECTOR &local_cell_vector)
    {
      const auto& fe_values = cdc.GetFEValuesState();
      unsigned int n_dofs_per_cell = cdc.GetNDoFsPerCell();
      unsigned int n_q_points = cdc.GetNQPoints();
\end{lstlisting}
 The \texttt{CellDataContainer} contains all the information we need to compute the cell-contributions (things like number of quadrature points, number of degrees of freedom on the cell, the \texttt{dealii::FEValues}-objects, etc.).
  \begin{lstlisting}
     cdc.GetValuesState("last_newton_solution", uvalues);
     cdc.GetGradsState("last_newton_solution", ugrads);
 \end{lstlisting}
 In these two lines, we load the (gradient of the) solution of the last newton iteration in all the quadrature points on the cell into the two vectors \texttt{uvalues} and \texttt{ugrads}.

In the next lines, we finally implement our concrete PDE. Here, aim to provide 
an interface which is closely related to the weak formulation which 
has to be derived by the user. Thus,
we loop over the quadrature points and the number of DoFs of the cell. In the innermost loop, we evaluate the weak form tested with $\phi_{i,h}$ (represented by \texttt{phi\_i\_grads\_v}) on the cell with the help of the previously selected quadrature formula. 
 \begin{lstlisting}
    for(unsigned int q_point = 0; q_point < n_q_points; q_point++)
    {
      //initialize phi_i_grads_v etc.
      Tensor<2,2> vgrads;
      vgrads[0]= ugrads[q_point][0]; 
      vgrads[1]= ugrads[q_point][1];
      for (unsigned int i=0;i<n_dofs_per_cell;i++)
        local_cell_vector(i) += ( _nu 
                                * vgrads * phi_i_grads_v 
                                + //all the other terms)
                                * state_fe_values.JxW(q_point);
    }
\end{lstlisting}
\dope{} takes now care of the distribution of the local cell contributions into  the vector of the residual \eqref{eq:residual_vector}. The methods \texttt{CellRightHandSide} and \texttt{CellMatrix} follow the same structure as 
\texttt{CellEquation}, so we do not go into the details here.

\subsubsection{\texttt{localfunctional.h}}
The implementation of the boundary-functional $J$ follows the outline of the computation of the cell-residual in \texttt{localpde.h}. However, there are two main differences. First, we do not have to specify the method \texttt{CellEquation}, but \texttt{BoundaryValue} instead. This is due to the fact that we consider a boundary-functional in this example, and so we have to specify the value of the functional not cell-wise but face wise (for the faces that lie on $\Gamma_{circ}$). The second difference is that as we only want to evaluate a functional and not a residual-vector, we just loop over the quadrature points and not over the degrees of freedom on the face.
\begin{lstlisting}
 double
 BoundaryValue(const FaceDataContainer<2>& fdc)
  {
    //same as before, get n_q_points etc.
    fdc.GetFaceValuesState("state", ufacevalues);
    fdc.GetFaceGradsState("state", ufacegrads);
\end{lstlisting}
Notice that we get here the values of the computed solution by giving the string "state".
\begin{lstlisting}
    for(unsigned int q_point = 0; q_point < n_q_points; q_point++)
      erg += -1. * ( _nu * ufacegrads[q_point][0]
                * state_fe_face_values.normal_vector(q_point)
                + //other terms)        
                * state_fe_face_values.JxW(q_point);
    return erg;
  }
\end{lstlisting}

>>>>

%\newcommand{\todo}[1]{\textbf{\textsc{\textcolor{black}{TODO: #1}}}}
\newcommand{\todo}[1]{\textbf{\textsc{\textcolor{blue}{TODO: #1}}}}
\newcommand{\todocg}[1]{\textbf{\textsc{\textcolor{red}{TODO: #1\textasciitilde cg}}}}
%\newcommand{\mymarginpar}[1]{\marginpar{\textcolor{red}{\scriptsize{#1}}}}
\newcommand{\mymarginpar}[1]{\marginpar{\textcolor{red}{\normalsize{#1}}}}
%
\begin{document}
\markboth{C. Goll, T. Wick, and W. Wollner}{DOpElib:  A Goal Oriented Software Library}

@@

\begin{abstract}
\todo{Computing}
In this article, we describe the software library 
\textit{Differential Equations and Optimization Environment} (\dope{}).
The main feature of \dope{} is that it provides a unified interface to high level algorithms 
such as time-stepping methods, nonlinear solvers and optimization routines. This structure ensures 
that first of all the user is required to write only those sections of code that are specific to 
the considered problem. Second, the exchange of parts of the used routines is possible 
with only a few lines of code to change instead of large reimplementations.
The article illustrates the design principles and various features
of \dope{} and provides some 
numerical results as demonstration for the versatility of the software.
\end{abstract}

\begin{bottomstuff}
Author's adresses: C. Goll, Institut f\"ur Angewandte Mathematik,
Universit\"at Heidelberg;
T. Wick, The Institute for Computational Engineering and Sciences, The
University of Texas at Austin;
W. Wollner, Fachbereich Mathematik, Universit\"at Hamburg.
\end{bottomstuff}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exist a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \cite{dealnew}, dune \cite{dune}, 
FEniCS \cite{fenics}
or commercial solvers as for instance ANSYS (with its Fluent package),
COMSOL, or simufact forming.
Finally, a lot of work is (still) done in Matlab.
In fact, these libraries provide very useful tools 
for a huge variety of applications.

In many cases, the user is confronted with 
certain bottlenecks. On the one hand, 
in open-source projects for academic research, 
standard high-level routines 
which are necessary to solve even simplest PDEs,
such as linear and nonlinear solvers or time-stepping schemes must be implemented 
by the users themselves. However, such routines are, at large,
problem-independent and could be provided as standard tools. 
As our guiding design principle, we note that in many 
numerical studies, the main task is the computation of some  
target functional, e.g., error norms,
point values, or certain mean-values as drag or lift coefficients.
As a second motivating principle, 
there is nowadays an increasing interest in computing 
optimization problems in which the state equation is governed 
by a PDE. Such optimization problems comprise optimal control,
parameter estimation, inverse problems or optimal experimental design. 
To do this in complex applications usually requires tremendous work already
on the PDE itself before considering optimization of the processes modeled 
by the PDE. Thus it is advantageous to have an environment that allows to 
focus on the development of an appropriate code for the PDE at first but 
asserts due to the required interface, that the code written for the PDE
is then reusable for the application of state-of-the-art optimization 
routines.
So far, there exist a very limited number of open source packages
which fulfill the requirements listed above; 
sucessful in-house code examples are Gascoigne \cite{gascoigne}
and RoDoBo \cite{rodobo}. 
Like DOpElib, both have been initiated as well in 
the Numerical Analysis Group in Heidelberg.

@@

The presented toolkit \dope{}: 
\textit{Differential Equations and Optimization Environment} \cite{dope}
fills this gap. We present a software
to solve both pure PDE problems as well as optimization problems 
constrained by a PDE. 
Its main feature is to give a unified interface to high level algorithms such as 
time-stepping methods, nonlinear solvers and optimization routines. 
\dope{} is designed such a way that the user only needs to write those parts
of the code that are problem dependent while all invariant 
parts of the algorithms
are reusable without any need for further coding.
In particular, the user is enabled to switch between various different 
algorithms without the need to rewrite the problem dependent code, 
though obviously he or she will
have to replace the algorithm object with an other one. 
This replacement can be done by replacing the appropriate object at only
one point in the code.

@@

The project was initially initiated at the University of Heidelberg in 2010 and  
at the present stage the following key features are supported by the library
\begin{itemize}
\item Solving of nonlinear stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various, easily exchangeable, 
  time-stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item Use of all finite elements of from \deal{} including hp-support.
\item Some examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Line search and trust region newton algorithms for the 
   solution of optimization problems with PDE constraints by elimination of 
   the state variable.
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other control and state constraints.
\item Several examples showing how to solve various kinds of optimization 
  problems involving PDE constraints.
\item Goal-oriented mesh adaptation with the Dual Weighted Residual method and
  standard residual based error estimators.
\item Different spatial triangulations for control and state variables without
  the need to reimplement the user dependent code when switching from one common 
  triangulation to different ones.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
This library is designed to allow easy implementation for the numerical solution 
of problems involving partial differential equations (PDEs). 
In particular, we remove all tedious routine implementation needed in 
\deal{} by introducing appropriate classes that take care of all the 
problem independent implementation while leaving only the problem specific 
parts to the user. To show how this is designed we will, in the sequel, 
describe the reasoning first for a simple stationary PDE in 
Section~\ref{subsubsec:stationary problems} and then show
how this idea extends to non stationary problems in 
Section~\ref{sec:timedep} as well as to optimization
problems with PDE constraints in Section~\ref{sec:opt}.

\subsection{Solving of Stationary PDEs}
\label{subsubsec:stationary problems}
For the start we consider a prototypical problem given by the solution of 
a stationary PDE, which we assume to be given in weak form.
In abstract notation it reads; find some $u$ in an appropriate set
such that
\begin{align}\label{eq:prototype_weak}
a(u)(\phi) = (f,\phi) \quad \forall \phi \in V,
\end{align}
with some suitable space $V$ for the test functions and a 
right hand side function $f$.
To illustrate this further, we take the stationary 
incompressible Navier-Stokes equations in a domain $\Omega \subset \R^d$, 
i.e., $u = (v,p)$ and $\phi = (\phi_u,\phi_p)$ then
\begin{align}\label{eq:ns}
a(u)(\phi) = \nu(\nabla v, \nabla\phi_u) + (v \nabla v,\phi_u) + (p, \nabla \cdot \phi_u) + (\nabla \cdot v ,\phi_p)
\end{align}
for all $\phi \in V = H^1_0(\Omega;\R^d) \times L^2_0(\Omega)$.
Here $(\cdot,\cdot)$ denotes the $L^2$-inner-product.
The solution will be searched for in $V + (u_d,0)$ where $u_d$ is some 
continuation of the given Dirichlet boundary data.
Then typically we do not only want to calculate $u$ but more likely we are 
interested in one (or several) functionals depending on the solution, i.e., some numbers $J$.

In order to solve this problem, i.e., calculate a sequence of values $J(u_h)$ approximating $J(u)$,
we need to discretize 
the equation. In \dope, we are using finite elements to do this. This means we need to pick a 
subdivision $\mathcal T_h$, for instance into quadrilaterals,
of the domain $\Omega$ and a finite element. In the concrete case, we could, 
for instance, take the $\mathcal Q_2/\mathcal Q_1$ Taylor-Hood element.
As it is clear, that the initialization of the unknowns, the local shape 
functions, removal of hanging node constraints, etc. always follows the 
same procedure, we define an object which takes the finite element (FE)
and the triangulation and takes care of all other things related to
handling the finite element space $V_h$. We will refer, to 
this as the \texttt{SpaceTimeHandler}. Then, it is clear that the simplest 
form of the constructor should have the following interface:
\begin{lstlisting}
  template<typename TRIANGULATION & triag, typename FE & fe>
      void SpaceTimeHandler(Triangulation &, FE);
\end{lstlisting}
Next, one needs to solve the discrete equation
\begin{align}\label{eq:discrete_equation}
a(u_h)(\phi_h) = (f,\phi_h) \quad \forall \phi \in V_h.
\end{align}
To this end, one can employ a newton type method with some globalization 
like line-search. To do so, one needs to calculate the residual of the 
equation, as well as to solve linearized equations for the calculation of 
the newton method and then employing some kind of globalization strategy.
This again will usually require the evaluation of the residual.

@@

\begin{remark}
In fact, the above constructors are not exactly what is implemented in \dope{}.
This is due to the fact, that for instance 
there are several different sparsity patterns available in 
\deal{} which the user should be able to choose. Thus it needs to be 
given to the \texttt{SpaceTimeHandler}. However, to keep the 
presentation clear we refrained from including these details above and in
what follows.
\end{remark}

In the next section we will now sketch the way a user needs to implement the
user-dependent data as the local cell integrals and the initialization of the
objects described above.
\subsubsection{Problem dependent implementation of stationary PDE problems}
To solve a stationary PDE problem, the user has to provide some code which can roughly be divided into the following main sections:
\begin{description}
\item[Discretization] Specify what kind of triangulation:
in most cases this is the
finite element and quadrature rules. Furthermore, other
things might be adjusted as well.
\item[Problem-data] Define the weak form, the corresponding matrices, the
  functionals $J$ as well as the boundary conditions. With exception of the boundary conditions, we have to specify all this only cell-wise, \dope{} handles the rest for us.
\item[Solvers] Choice of a (non-)linear solver. 
\item[Additional features] ParameterReader to read in the run-time parameters
  such as material coefficients or adjustments for the solvers like tolerance
  for Newton's method. 
\end{description}

We describe now a program designed to compute a value $J$, where $u=(v,p)$ is the solution of the Navier-Stokes equations \eqref{eq:ns} on the domain $\Omega$ as seen in figure~\ref{fig:example_ns} with  $f=0$, and $J$ is the drag-force of the enclosed circle. We assume given Dirichlet data for the velocity $v$ on the boundary $\Gamma_{D}=\Gamma_{in}\cup\Gamma_{wall}\cup\Gamma_{circ}$. This is the Benchmark-Case 2D-1 from \cite{TuSchae96}, see there for further details.
\begin{figure}[h]
\centering
\resizebox{0.5\textwidth}{!}{\input{Pictures/example_ns}}
\caption{Flow around a cylinder with 
circle-center $C=(0.2,0.2)$ and radius $r=0.05$.}
\label{fig:example_ns}
\end{figure}

\todocg{Haengt
  von style-file ab: textbf finde ich besser als textit, haengt aber vom stylefile ab.}
We split the program into the parts \texttt{main.cc}, \texttt{functionals.h} and \texttt{localpde.h}. The \texttt{main.cc}-file is responsible for the declaration of various objects which have mostly to do with the \textit{discretization}, the \textit{solvers} and the general \textit{program sequence}, whereas the other two contain most of the problem-data -- especially the cell-wise contributions of the weak form, the right hand side and the functional. We give a brief explanation on every file in the following.

@@

\end{lstlisting}
Next, we incorporate the Dirichlet data, which is given by the class
\texttt{DirichleData}. Then we have to tell the \texttt{prob\_container} for
which components and boundaries we prescribe the Dirichlet conditions. This is
done via a boundary-color in a so-called component mask 
(here, $0$ denotes $\Gamma_D$).
\begin{lstlisting}

@@

  typedef StatPDEProblem<NEWTON> PDE;

\end{lstlisting}
Here, \texttt{NEWTON} itself is an abbreviation of
\begin{lstlisting}

  typedef NewtonSolver<LINSOLVE> NEWTON;

\end{lstlisting}
The  template parameter \texttt{LINSOLVE} describes which linear solver we use in this example. We opted for a direct solver, so: 
\begin{lstlisting}

  typedef DirectLinearSolver<MATRIX, VECTOR> LINSOLVE;

@@

  PDE pde(prob_container, pr,
                    quad_formula, face_quad_formula);

\end{lstlisting}
Finally, we compute the functional $J$ by calling:
\begin{lstlisting}

  pde.ComputeReducedFunctionals();

@@

Let $N$ be the number of degrees of freedom of our discretization and $\Set{\phi_{h}^{i}|1\leq i \leq N}$ be a given FE-Basis. As stated in subsection~\ref{subsubsec:stationary problems}, to solve the discrete equation \eqref{eq:discrete_equation} we need the residual, i.e. the vector
\begin{align}\label{eq:discrete_residual}
\left(a(u_h)(\phi_{h}^{i})-(f,\phi_{h}^{i})\right)_{i=1}^N.
\end{align}
The corresponding Jacobian can be defined analogously.
Now all we have to supply here are the cell-wise contributions to these terms, the \texttt{Integrator} takes care of assembling \eqref{eq:discrete_residual} and \eqref{eq:discrete_matrix}.

To this end, we define a class \texttt{LocalPDE} which provides the three methods
\texttt{CellEquation}, \texttt{CellRightHandSide} and \texttt{CellMatrix}. 
The splitting of these terms is similar to Gascoigne
\cite{gascoigne} whereas the content (i.e., the 
implementation itself) of these
terms works by and large as in \deal{}.

We show the basic structure of these methods using the example of \texttt{CellEquation}.

\begin{lstlisting}
  void CellEquation(const CellDataContainer &cdc,
                    VECTOR &local_cell_vector)
    {
      const auto& fe_values = cdc.GetFEValuesState();
      unsigned int n_dofs_per_cell = cdc.GetNDoFsPerCell();
      unsigned int n_q_points = cdc.GetNQPoints();
\end{lstlisting}
 The \texttt{CellDataContainer} contains all the information we need to compute the cell-contributions (things like number of quadrature points, number of degrees of freedom on the cell, the \texttt{dealii::FEValues}-objects, etc.).
  \begin{lstlisting}
     cdc.GetValuesState("last_newton_solution", uvalues);
     cdc.GetGradsState("last_newton_solution", ugrads);
 \end{lstlisting}
 In these two lines, we load the (gradient of the) solution of the last newton iteration in all the quadrature points on the cell into the two vectors \texttt{uvalues} and \texttt{ugrads}.

In the next lines, we finally implement our concrete PDE. Here, aim to provide 
an interface which is closely related to the weak formulation which 
has to be derived by the user. Thus,
we loop over the quadrature points and the number of DoFs of the cell. In the innermost loop, we evaluate the weak form tested with $\phi_{h}^i$ (represented by \texttt{phi\_i\_grads\_v}) on the cell with the help of the previously selected quadrature formula. 
 \begin{lstlisting}
    for(unsigned int q_point = 0; q_point < n_q_points; q_point++)
    {
      //initialize phi_i_grads_v etc.
      Tensor<2,2> vgrads;
      vgrads[0]= ugrads[q_point][0]; 
      vgrads[1]= ugrads[q_point][1];
      for (unsigned int i=0;i<n_dofs_per_cell;i++)
        local_cell_vector(i) += ( _nu 
                                * vgrads * phi_i_grads_v 
                                + //all the other terms)
                                * state_fe_values.JxW(q_point);
    }
\end{lstlisting}
\dope{} takes now care of the distribution of the local cell contributions into  the vector of the residual \eqref{eq:residual_vector}. The methods \texttt{CellRightHandSide} and \texttt{CellMatrix} follow the same structure as 
\texttt{CellEquation}, so we do not go into the details here.

\subsubsection{\texttt{localfunctional.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{localpde.h}. 
\todo{CG and WW: Pruefen, ob der folgende Absatz nun gut ist:}
The main difference is that as we only want to evaluate a functional and not a
residual-vector, we just loop over the quadrature points and not over the
degrees of freedom on the face. Another deviation, special to the case at
hand, is that we do not have to specify the method \texttt{CellEquation}, but
\texttt{BoundaryValue} instead because $J$ is just number to be computed. This is due to the fact that we consider a boundary-functional in this example, and so we have to specify the value of the functional not cell-wise but face wise (for the faces that lie on $\Gamma_{circ}$).

%However, there are two main differences. First, we do not have to specify the method \texttt{CellEquation}, but \texttt{BoundaryValue} instead. This is due to the fact that we consider a boundary-functional in this example, and so we have to specify the value of the functional not cell-wise but face wise (for the faces that lie on $\Gamma_{circ}$). The second difference is that as we only want to evaluate a functional and not a residual-vector, we just loop over the quadrature points and not over the degrees of freedom on the face.
\begin{lstlisting}
 double
 BoundaryValue(const FaceDataContainer& fdc)
  {
    //same as before, get n_q_points etc.
    fdc.GetFaceValuesState("state", ufacevalues);
    fdc.GetFaceGradsState("state", ufacegrads);
\end{lstlisting}
Notice that we get here the values of the computed solution by giving the string "state".
\begin{lstlisting}
    for(unsigned int q_point = 0; q_point < n_q_points; q_point++)
      erg += -1. * ( _nu * ufacegrads[q_point][0]
                * state_fe_face_values.normal_vector(q_point)
                + //other terms)        
                * state_fe_face_values.JxW(q_point);
    return erg;
  }
\end{lstlisting}
